Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-07-08 tex/scam/summary.tex
added optimization subsection

\subsection{Additive faithfulness}

The starting point for our approach is the observation that least squares
nonparametric estimation under convexity constraints is equivalent to
a finite dimensional quadratic program.  Specifically, the infinite
dimensional optimization 
\begin{align}
\begin{split}
\text{minimize} & \quad \sum_{i=1}^n (Y_i - f(\x_i))^2 \\
\text{subject to} &  \quad f:\reals^p\rightarrow\reals\ \text{is
  convex}
\end{split}
\end{align}
is equivalent to the finite dimensional quadratic
program 
\begin{align}
\begin{split}
\text{minimize}_{f, \beta} & \;\; \sum_{i=1}^n (Y_i - f_i)^2 \\
\text{subject to} & \;\; f_j \geq f_i + \beta_i^T (\x_j-\x_i),\; \text{for
    all $i,j$}.
\end{split}
\end{align}
See \cite{Boyd04}, Section 6.5.5.
Here $f_i$ is the estimated function value $f(\x_i)$, and the vectors
$\beta_i \in \reals^d$ represent supporting hyperplanes to the
epigraph of $f$.  Importantly, this finite dimensional quadratic program does
not have tuning parameters for smoothing the function.

@@

\begin{theorem}
Suppose $p(\mathbf{x})$ is a positive
density on $C=[0,1]^p$ and satisfies the boundary-points
condition. Suppose that $f$ is convex and twice-differentiable.
and that $\partial_{x_k} f$, $\partial_{x_k} p( \mathbf{x}_{-k}
\given x_k )$, and $\partial_{x_k}^2 p( \mathbf{x}_{-k} \given x_k)$
are all continuous as functions on $C$.\\
Define
\begin{equation}
\{ f^*_k \}_{k=1}^p,\mu^* = \arg\min \Big \{
\E\big( f(X) - \mu - \sum_{k=1}^s f_k(X_k) \big)^2 \,:\, f_k \in
\mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation} where $\mathcal{C}^1$ is the set of univariate convex
functions, and
\begin{equation}
g^*_k = \arg\min \Big\{ \E\big( f(X) -
\sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 \,:\, g_k \in \mh
\mathcal{C}^1, \E g_k(X_k) = 0 \Big\},
\end{equation}
with $\mh{}\mathcal{C}^1$ denoting the set of univariate concave
function.  Then, $f^*_k = 0$ and $g^*_k = 0$ implies that $f$ does not
depend on $x_k$, i.e., $\partial_{x_k} f(\mathbf{x}) = 0$ with
probability one.
\end{theorem}

This result suggests a two-stage screening
procedure for variable selection. In the first stage we fit a sparse \emph{convex}
additive model.  In the second stage we
fit a concave function $\hat g_k$ to the residual for each variable
having a zero convex component $\hat f_k$.  If both $\hat f_k = 0$ and
$\hat g_k = 0$, we remove variable $x_k$.  
Our next results concern the optimizations involved, and their finite
sample statistical performance.

@@

>>>>

\subsection{Faithful screening}

The starting point for our approach is the observation that least squares
nonparametric estimation under convexity constraints is equivalent to
a finite dimensional quadratic program.  Specifically, the infinite
dimensional optimization 
\begin{align}
\begin{split}
\text{minimize} & \quad \sum_{i=1}^n (Y_i - f(\x_i))^2 \\
\text{subject to} &  \quad f:\reals^p\rightarrow\reals\ \text{is
  convex}
\end{split}
\end{align}
is equivalent to the finite dimensional quadratic
program 
\begin{align}
\label{eq:convreg}
\begin{split}
\text{minimize}_{f, \beta} & \;\; \sum_{i=1}^n (Y_i - f_i)^2 \\
\text{subject to} & \;\; f_j \geq f_i + \beta_i^T (\x_j-\x_i),\; \text{for
    all $i,j$}.
\end{split}
\end{align}
See \cite{Boyd04}, Section 6.5.5.
Here $f_i$ is the estimated function value $f(\x_i)$, and the vectors
$\beta_i \in \reals^d$ represent supporting hyperplanes to the
epigraph of $f$.  Importantly, this finite dimensional quadratic program does
not have tuning parameters for smoothing the function.

@@

\begin{theorem}
Suppose $p(\mathbf{x})$ is a positive
density on $C=[0,1]^p$ and satisfies the boundary-points
condition. Suppose that $f$ is convex and twice-differentiable.
and that $\partial_{x_k} f$, $\partial_{x_k} p( \mathbf{x}_{-k}
\given x_k )$, and $\partial_{x_k}^2 p( \mathbf{x}_{-k} \given x_k)$
are all continuous as functions on $C$.\\
Define
\begin{equation}
\{ f^*_k \}_{k=1}^p,\mu^* = \arg\min \Big \{
\E\big( f(X) - \mu - \sum_{k=1}^s f_k(X_k) \big)^2 \,:\, f_k \in
\mathcal{C}^1, \, \E f_k(X_k) = 0 \Big \}
\end{equation} where $\mathcal{C}^1$ is the set of univariate convex
functions, and
\begin{equation}
g^*_k = \arg\min \Big\{ \E\big( f(X) -
\sum_{k' \neq k} f^*_{k'}(X_{k'}) - g_k \big)^2 \,:\, g_k \in \mh
\mathcal{C}^1, \E g_k(X_k) = 0 \Big\},
\end{equation}
with $\mh{}\mathcal{C}^1$ denoting the set of univariate concave
functions.  Then $f^*_k = 0$ and $g^*_k = 0$ implies that $f$ does not
depend on $x_k$, i.e., $\partial_{x_k} f(\mathbf{x}) = 0$ with
probability one.
\end{theorem}

This result naturally suggests a two-stage screening
procedure for variable selection. In the first stage we fit a sparse convex
additive model.  In the second stage we
fit a concave function $\hat g_k$ to the residual for each variable
having a zero convex component $\hat f_k$.  If both $\hat f_k = 0$ and
$\hat g_k = 0$, we can safely discard variable $x_k$.  
Our next results concern the required optimizations, and their finite
sample statistical performance.

@@

In Section~\ref{sec:optimization} we present optimization algorithms for the additive convex regression stage.
The convex constraints for the additive functions, analogous to 
the multivariate constraints \eqref{eq:convreg},
are  that each component $f_{k}(\cdot)$ 
can be represented by its supporting hyperplanes, i.e.,
\begin{equation}
      f_{ki'} \geq f_{ki} + \beta_{ki}(x_{ki'}-x_{ki}) \quad \text{for
        all $i,i'$}
\end{equation}
where $f_{ki}\coloneqq f_{k}(x_{ki})$ and $\beta_{ki}$ is the
subgradient at point $x_{ki}$. While this apparently requires $O(n^2
p)$ equations to impose the supporting hyperplane constraints, 
in fact, only $O(np)$ constraints suffice.  This is because univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to a reduced quadratic
program with $O(np)$ variables and $O(np)$ constraints.

Directly applying a QP solver to this optimization is still computationally
expensive for relatively large
$n$ and $p$.  We thus develop a block
coordinate descent method, where in each step we solve a sparse
QP involving $O(n)$ variables and $O(n)$ constraints.  This 
is efficiently solved using optimization packages 
such as {\sc mosek}.  The details of these optimizations
are given in Section~\ref{sec:optimization}.
